Open archaeology, open source?

### 2024-02-02 analysis/_01-introduction.qmd
Mostly minor but necessary fixes to code and text
Provided a little explanation for the manual licenses file, included a missing value to the licenses file, corrected some citations, added section slugs and encoded links to sections, removed various completed or unnecessary todos and notes, removed dendrogram, verified and updated characterization of clusters, fixed references to graph objects
Updates to the text and general touch ups.
Moved stuff on git and github to a new background subsection, moved data caveats from conclusion to methods, updated some interpretations in section 6, update i

# Introduction

@@

<!-- Might be getting a bit ahead of ourselves here. --JR -->That being said, archaeologists are prolific software developers. This is made evident by open-archaeo, a comprehensive list of open source archaeological software and resources that indexes <!-- TODO: encode -->XXX records, as of the time of this article's publication. But beyond simply making their code available on the web, do archaeologists also implement social strategies to advance open source ideals? Does archaeological open source actually help achieve greater transparency, sustainability, and community participation? And if not, what does it actually achieve?

>>>>

# Introduction {#sec-introduction}

@@

<!-- Might be getting a bit ahead of ourselves here. --JR -->That being said, archaeologists are prolific software developers. This is made evident by open-archaeo, a comprehensive list of open source archaeological software and resources that indexes 501 records, as of the time of this article's publication. But beyond simply making their code available on the web, do archaeologists also implement social strategies to advance open source ideals? Does archaeological open source actually help achieve greater transparency, sustainability, and community participation? And if not, what does it actually achieve?

### 2024-02-02 analysis/_02-background.qmd
Mostly minor but necessary fixes to code and text
Provided a little explanation for the manual licenses file, included a missing value to the licenses file, corrected some citations, added section slugs and encoded links to sections, removed various completed or unnecessary todos and notes, removed dendrogram, verified and updated characterization of clusters, fixed references to graph objects
Updates to the text and general touch ups.
Moved stuff on git and github to a new background subsection, moved data caveats from conclusion to methods, updated some interpretations in section 6, update i

# Open science and open source

@@

The adoption of open source development models among archaeologists is generally informed by the broader open science movement, which is motivated by a genuine desire to facilitate novel research opportunities, to make participation in scientific research more equitable, to reclaim science as a public good, and to enhance the means of validating findings. However, the predominant concern with implementing best tools to use, adopting optimal data processing pipelines, and tying into global, web-based infrastructures, protocols and standards [cf. @kansa2014; @kintigh2015; @roosevelt2015] distract from fundamental tensions and contradictions regarding the actual value of working in the open. For instance, Faniel et al. [-@faniel2013 299-301], Atici et al. [-@atici2013 676-677], Huggett [-@huggett2018; -@huggett2022], @sobotkova2018, @opitz2021, @haciguzeller2021, and @batist2023a demonstrate that to make the reuse of archaeological data feasible and useful in a practical sense, it is necessary to re-introduce social friction that these infrastructures are designed to eliminate. In other words, the pressures and circumstances of being an archaeologist and doing archaeological research assert themselves when attempting to make practical use of these infrastructures, and therefore must be accounted for in their design and implementation. Here we aim to identify similar sources of dissonance with regards to the promise, potential, and actual implementation of open source software development models among archaeologists.

>>>>

# Open science and open source {#sec-background}

@@

The adoption of open source development models among archaeologists is generally informed by the broader open science movement, which is motivated by a genuine desire to facilitate novel research opportunities, to make participation in scientific research more equitable, to reclaim science as a public good, and to enhance the means of validating findings. However, the predominant concern with implementing best tools to use, adopting optimal data processing pipelines, and tying into global, web-based infrastructures, protocols and standards [cf. @kansa2014; @kintigh2015; @roosevelt2015] distract from fundamental tensions and contradictions regarding the actual value of working in the open. For instance, Faniel et al. [-@faniel2013 299-301], Atici et al. [-@atici2013 676-677], Huggett [-@huggett2018; -@huggett2022], @sobotkova2018, @opitz2021, @haciguzeller2021, and @batist2023a demonstrate that to make the reuse of archaeological data feasible and useful in a practical sense, it is necessary to re-introduce social friction that these infrastructures are designed to eliminate. In other words, the pressures and circumstances of being an archaeologist and doing archaeological research assert themselves when attempting to make practical use of these infrastructures, and therefore must be accounted for in their design and implementation. In this paper we aim to identify similar sources of dissonance with regards to the promise, potential, and actual implementation of open source software development models among archaeologists.

## Open Source Infrastructure

Open source is an inherently internet-based development model and is supported by technical infrastructures which facilitate global distribution of labour and code.
Here we provide a brief overview of key technologies that archaeologists have come to rely on as they develop open source software.

Chief among these is git, a protocol designed to facilitate open and distributed contributions to a common code base.
It operates by providing mechanisms for synchronizing communal, web-based public repositories with local iterations stored on contributors' private workstations.
Contributors who volunteer or are assigned to develop, inspect, or revise a specific aspect of a code base download a copy of the public repository into their own work environment, create a fork in which they apply their modifications, and then request that their fork be merged into the central code base.
After a public repository's maintainers decide to merge the proposed changes into the communal code base, other developers may use git to download these changes while keeping their own independent forks intact.

Git is also designed to facilitate code review and version control.
All modifications are tracked as "diffs", which highlight additions or deletions to the code base, including changes within individual files.
Typically, a contributor will group a series of changes into a more comprehensive "commit" based on a specific task or part of a workflow.
Commits are always accompanied by a message, in which the contributor (ideally) describes the reason and context for the changes included in the commit.
Moreover, git assigns each commit a unique identifier and identifies the contributor by name and email address to ensure some degree of public accountability.

Forges, or collaborative software development platforms like GitHub, GitLab and Codeberg are designed to facilitate open source software development by hosting public git repositories.
However, they also support common software developer and project management practices, such as issue and bug tracking, code-commenting, task management, identity and permissions management, web publishing, vulnerability detection, creation and maintenance of metadata, and financial sponsorship.
These platforms also implement standard social media functions, like the ability to follow projects and individual users to receive updates on their activities, "star" certain repositories as favourites, and maintain a public-facing profile that includes personally-identifying information (e.g. profile picture, username, real name, employer or affiliation), references to all public activity on the platform, and links to the user's other social media profiles.
Code-sharing platforms thus serve as comprehensive developer portfolios and community networking resources.
While these additional features are meant to complement and enhance the experience of contributing to open source projects, they are not actually part of the git protocol.
See @tbl-glossary for a glossary of terminology used to describe many of these concepts and features, and which we use throughout this paper.

```{r tbl-glossary}
#| tbl-cap: Glossary of git and GitHub terminology
tribble(
  ~term, ~definition,
  "Comment", "On GitHub, text post attached to an issue, including the first one that describes the issue",
  "Commit", "Set of changes (addition, alteration, or deletion) to files in a repository that has been recorded by git as one entry in its log",
  "Commit access", "Ability to make changes to a repository directly, without making a pull request",
  "Contributor", "User that has made at least one commit to a specified repository",
  "Follow", "Add activity by another user to a user's timeline",
  "Forge", "Web-based platform for hosting, distributing and facilitating collaboration on version controlled computer code, e.g. GitHub, GitLab, Codeberg",
  "Fork", "Copy of a repository owned by another user; forking is a prerequisite to making a pull request",
  "git", "Open source version control software",
  "GitHub", "Commercial platform that freely hosts git repositories and provides extended collaboration and social networking features, such as pull requests, issues and stars",
  "GitLab", "Open source alternative to GitHub",
  "CodeBerg", "Open source alternative to GitHub",
  "Issue", "Feature of GitHub that records and tracks a bug report, feature request or other suggestion in a repository",
  "Maintainer", "Individual that has overall control of a repository, generally assumed to be its primary contributor. Repositories can have multiple users with commit access in addition to the maintainer.",
  "Merge", "Accept a pull request and incorporate its changes into a repository.",
  "Organization", "Entity representing a group of users, which can also own repositories",
  "Pull request", "Mechanism by which users that don't have commit access to a repository can contribute to it. The repository's maintainer or another user with commit access must decide whether to merge (accept) the changes, or decline them.",
  "Repository", "Individual project that uses git for version control. Can include a mix of different types of files.",
  "Star", "GitHub's version of a 'like', applied by users to a repository.",
  "Timeline", "Chronological feed of GitHub activity from repositories a user has starred and other users they follow. Also includes repositories that a user is not following if they are 'trending' or determined relevant by GitHub's algorithm.",
  "User", "On GitHub, an individual with an account that can own repositories",
  "Version control", "System for tracking changes (additions, alterations, or deletions) in a set of files, typically but not exclusively computer code"
) |>
  rename_with(tools::toTitleCase) |>
  tt() |>
  format_tt(2, markdown = TRUE) |>
  style_tt(j = 1, italic = TRUE)
```

### 2024-02-02 analysis/_03-data_methodology.qmd
Updates to the text and general touch ups.
Moved stuff on git and github to a new background subsection, moved data caveats from conclusion to methods, updated some interpretations in section 6, update interpretation based on cumulative growth figure.

```{r tbl-glossary}
#| tbl-cap: Glossary of git and GitHub terminology
tribble(
  ~term, ~definition,
  "Comment", "On GitHub, text post attached to an issue, including the first one that describes the issue",
  "Commit", "Set of changes (addition, alteration, or deletion) to files in a repository that has been recorded by git as one entry in its log",
  "Commit access", "Ability to make changes to a repository directly, without making a pull request",
  "Contributor", "User that has made at least one commit to a specified repository",
  "Follow", "Add activity by another user to a user's timeline",
  "Forge", "Web-based platform for hosting, distributing and facilitating collaboration on version controlled computer code, e.g. GitHub, GitLab, Codeberg",
  "Fork", "Copy of a repository owned by another user; forking is a prerequisite to making a pull request",
  "git", "Open source version control software",
  "GitHub", "Commercial platform that freely hosts git repositories and provides extended collaboration and social networking features, such as pull requests, issues and stars",
  "GitLab", "Open source alternative to GitHub",
  "CodeBerg", "Open source alternative to GitHub",
  "Issue", "Feature of GitHub that records and tracks a bug report, feature request or other suggestion in a repository",
  "Maintainer", "Individual that has overall control of a repository, generally assumed to be its primary contributor. Repositories can have multiple users with commit access in addition to the maintainer.",
  "Merge", "Accept a pull request and incorporate its changes into a repository.",
  "Organization", "Entity representing a group of users, which can also own repositories",
  "Pull request", "Mechanism by which users that don't have commit access to a repository can contribute to it. The repository's maintainer or another user with commit access must decide whether to merge (accept) the changes, or decline them.",
  "Repository", "Individual project that uses git for version control. Can include a mix of different types of files.",
  "Star", "GitHub's version of a 'like', applied by users to a repository.",
  "Timeline", "Chronological feed of GitHub activity from repositories a user has starred and other users they follow. Also includes repositories that a user is not following if they are 'trending' or determined relevant by GitHub's algorithm.",
  "User", "On GitHub, an individual with an account that can own repositories",
  "Version control", "System for tracking changes (additions, alterations, or deletions) in a set of files, typically but not exclusively computer code"
) |>
  rename_with(tools::toTitleCase) |>
  tt() |>
  format_tt(2, markdown = TRUE) |>
  style_tt(j = 1, italic = TRUE)
```

We compiled the dataset by browsing collaborative software development platforms such as GitHub, GitLab and Codeberg, which are designed to facilitate open source software development by hosting public git repositories.
Git is a protocol designed to facilitate open and distributed contributions to a common code base.
It operates by providing mechanisms for synchronizing communal, web-based public repositories with local iterations stored on contributors' private workstations.
Contributors who volunteer or are assigned to develop, inspect, or revise a specific aspect of a code base download a copy of the public repository into their own work environment, create a fork in which they apply their modifications, and then request that their fork be merged into the central code base.
After a public repository's maintainers decide to merge the proposed changes into the communal code base, other developers may use git to download these changes while keeping their own independent forks intact.

Git is also designed to facilitate code review and version control.
All modifications are tracked as "diffs", which highlight additions or deletions to the code base, including changes within individual files.
Typically, a contributor will group a series of changes into a more comprehensive "commit" based on a specific task or part of a workflow.
Commits are always accompanied by a message, in which the contributor (ideally) describes the reason and context for the changes included in the commit.
Moreover, git assigns each commit a unique identifier and identifies the contributor by name and email address to ensure some degree of public accountability.

In addition to serving as hosting providers for public git repositories, platforms like GitHub, GitLab and Codeberg support common software developer and project management practices, such as issue and bug tracking, code-commenting, task management, identity and permissions management, web publishing, vulnerability detection, creation and maintenance of metadata, and financial sponsorship.
They also implement standard social media functions, like the ability to follow projects and individual users to receive updates on their activities, "star" certain repositories as favourites, and maintain a public-facing profile that includes personally-identifying information (e.g. profile picture, username, real name, employer or affiliation), references to all public activity on the platform, and links to the user's other social media profiles.
Code-sharing platforms thus serve as comprehensive developer portfolios and community networking resources.
However, these additional features are meant to complement and enhance the experience of contributing to open source projects, but are not actually part of the git protocol.
See @tbl-glossary for a glossary of terminology used to describe many of these concepts and features, and which we use throughout this paper.

We rely heavily on the social networking features provided by GitHub, GitLab and Codeberg to maintain open-archaeo.
More specifically, we update open-archaeo by manually crawling through archaeologists' profiles on these platforms, as well as on other personal, professional, and institutional websites that describe and host additional archaeological software.
We supplement this quasi-systematic collection strategy with word-of-mouth contributions made by interested individuals who identified relevant work that we initially overlooked.

@@

We excluded `r n_items_removed` records from open-archaeo from this analysis because they did not fit our "made by archaeologists, for archaeologists" criteria.

@@

Where applicable, we obtained more detailed information about each repository's contents and contribution histories from the GitHub API (application programming interface).
We opted to only collect repository data from GitHub because it is the most popular forge platform used by open-archaeo projects (@tbl-forges).
This means that projects that do not use version control (`r percent(p_no_vc)` of the total), or host it elsewhere (`r percent(p_no_github)` of the total), are excluded from these parts of the analysis, though we were still able to perform an analysis of their contents and authorship from the data included in open-archaeo itself.
In total, we analysed data on `r n_repos` repositories, comprising `r n_commits` commits, `r n_issues` issues/pull requests, and `r n_comments` from `r n_contributors` distinct users, as well as repository metadata on programming languages used, licensing, stars and forks, and so on.

Taken together the open-archaeo directory and the supplemental data from the GitHub API provide a rich resource to explore the nature of collaborative software engineering in archaeology.
Here we employ exploratory data analysis [@tukey1977] to identify and characterise overall trends visible in this rich dataset.
In @sec-open-archaeology, our focus is on examining the general state of open source archaeological software and resource development.
In @sec-collaboration, we refine our analysis to examine development processes, with specific focus on collaborative experiences.
Finally, in @sec-network, we apply network analysis methods to investigate the formation of broader collaborative communities.
Our analyses combine to support our objectives of understanding what kinds of software and resources archaeologists are making, and how they create these tools in response to specific needs and use-cases, as afforded by offline social and professional connections, and within the context of an emerging community of practice.

>>>>

We compiled the dataset by browsing collaborative software development platforms, relying heavily on their social networking features.
More specifically, we update open-archaeo by manually crawling through archaeologists' profiles on these platforms, as well as on other personal, professional, and institutional websites that describe and host additional archaeological software.
We supplement this quasi-systematic collection strategy with word-of-mouth contributions made by interested individuals who identified relevant work that we initially overlooked.

@@

@@

Where applicable, we obtained more detailed information about each repository's contents and contribution histories from the GitHub API (application programming interface).
In total, we analysed data on `r n_repos` repositories, comprising `r n_commits` commits, `r n_issues` issues/pull requests, and `r n_comments` from `r n_contributors` distinct users, as well as repository metadata on programming languages used, licensing, stars and forks, and so on.

We opted to only collect repository data from GitHub because it is the most popular forge platform used by open-archaeo projects (@tbl-forges).
This means that projects that do not use version control (`r percent(p_no_vc)` of the total), or host it elsewhere (`r percent(p_no_github)` of the total), are excluded from these parts of the analysis, though we were still able to perform an analysis of their contents and authorship from the data included in open-archaeo itself.
We also exclude collaboration through offline or private channels, or forms of collaboration we do not know about. 
We did not directly observe or interview archaeological software developers, though our conclusions do draw heavily from our experience as members of that community ourselves. 
Our earliest data is from 2005 and our study can say little about collaborative software development in archaeology before this point, though we know there was a significant amount of it [@ducke2013 ; @whallon1972].

These caveats notwithstanding, the open-archaeo directory and the supplemental data from the GitHub API provide a rich resource to explore the nature of collaborative software engineering in archaeology.
Here we employ exploratory data analysis [@tukey1977] to identify and characterise overall trends visible in this rich dataset.
In @sec-open-archaeology, our focus is on examining the general state of open source archaeological software and resource development.
In @sec-collaboration, we refine our analysis to examine development processes, with specific focus on collaborative experiences.
Finally, in @sec-network, we apply network analysis methods to investigate the formation of broader collaborative communities.
Our analyses combine to support our objectives of understanding what kinds of software and resources archaeologists are making, and how they create these tools in response to specific needs and use-cases, as afforded by offline social and professional connections, and within the context of an emerging community of practice.

### 2024-02-02 analysis/_04-open_archaeology.qmd
Updates to the text and general touch ups.
Moved stuff on git and github to a new background subsection, moved data caveats from conclusion to methods, updated some interpretations in section 6, update interpretation based on cumulative growth figure.

As we can see, archaeologists have been using git from even before GitHub was launched in 2008.
But use of git really began to take off around 2014--2015, when we see an uptick in the rate of growth that has continued ever since.
<!-- TODO: Update and note plateau -->

Around this time we also see that GitHub starts being used to host documents and scripts.
This may represent a recognition of GitHub's ability to track things other than code, and a willingness to experiment with version control systems as a medium for disseminating work in an open and somewhat nerdy way.

>>>>

As we can see <!-- this is no longer evident in the figure -->, archaeologists have been using git from even before GitHub was launched in 2008.
But use of git really began to take off around 2014--2015, when we see an uptick in the rate of growth.
Around this time we also see that GitHub starts being used to host documents and scripts.
This may represent a recognition of GitHub's ability to track things other than code, and a willingness to experiment with version control systems as a medium for disseminating work in an open and somewhat nerdy way.

However, since around 2022, the number of new GitHub repositories has significantly plateaud, while the cumulative number of commits has continued to rise.
This may relate to a general emphasis on maintaining existing code and working on established projects, rather than spinning up new ones.

### 2024-02-02 analysis/_06-network.qmd
Updates to the text and general touch ups.
Moved stuff on git and github to a new background subsection, moved data caveats from conclusion to methods, updated some interpretations in section 6, update interpretation based on cumulative growth figure.

```{r fig-graph-repo-repo}
#| fig-cap: Graph of open archaeology repositories connected by common contributors. Darker edges indicate a great number of common contributors. Node colour indicates membership of the largest clusters according to the edge–betweenness method [@girvan2002]. Excludes isolate nodes.
g_repo_repo <- oarch_graph_repo_repo |>
  group_by(cluster) |>
  mutate(cluster_n = n()) |>
  ungroup() |>
  mutate(
    cluster_rank = dense_rank(desc(cluster_n)),
    cluster = if_else(cluster_rank <= 7, cluster, NA)
  ) |>
  ggraph(layout = "fr", niter = 5000, grid = "nogrid") +
  geom_edge_link(aes(alpha = n)) +
  geom_point_interactive(aes(x, y, colour = cluster, tooltip = repo)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  theme(panel.background = element_rect(fill = "white"))

@@

```{r fig-graph-user-user}
#| fig-cap: Graph of open archaeology users connected by contributions to common repositories. Darker edges indicate a great number of common repositories. Node colour indicates membership of the largest clusters according to the edge–betweenness method [@girvan2002]. Excludes isolate nodes.
g_user_user <- oarch_graph_user_user |>
  group_by(cluster) |>
  mutate(cluster_n = n()) |>
  ungroup() |>
  mutate(
    cluster_rank = dense_rank(desc(cluster_n)),
    cluster = if_else(cluster_rank <= 7, cluster, NA)
  ) |>
  ggraph(layout = "fr", niter = 5000, grid = "nogrid") +
  geom_edge_link(aes(alpha = n)) +
  geom_point_interactive(aes(x, y, colour = cluster, tooltip = user)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  theme(panel.background = element_rect(fill = "white"))

@@

Likewise, we applied the same edge-betweenness community detection method to identify clusters in the user-user network, which reveal collaborative groupings.
We again see a complementary primary core connected to several more peripheral clusters, which are internally-cohesive and exhibit few connections with other peripheral clusters.
The central core bridges all the peripheral clusters.
The central core is not uniform, and comprises several relatively discrete sub-groups representing collaborative sub-communities.
While these sub-groups are internally cohesive, they exhibit enough connections to other members of the central core so as to not be considered as separate or peripheral clusters.

@@

Despite being the minority language, Python packages tend to be more central than R packages, on average.
This might relate to the different ways in which R and Python developers call dependencies ???
<!-- @ZB: I think this is interesting, but I couldn't really think of a general observation to make here. Maybe you can? -->
<!-- I wonder what the median is? Also: cltk is a python script, and is the massive pink blob in the repo-repo and user-user networks, so that might skew things slightly -->

@@

<!-- Predictors of user centrality -->
Centrality to the user–user graph is weakly predicted by a user's overall rate of activity, as measured by their total number of contributions [@fig-user-betweenness-contribs]. 
We did not collect demographic data on users that appear in our dataset, but based on our own knowledge of the community we can observe that those highly central to the network tend to be employed in (junior) academic positions, or in a few cases in cultural heritage authorities, rather than specifically as research software engineers.
Such positions tend to not to actively reward or encourage software development, at least not on a par with more traditional academic outputs [@baxter2012], and are increasingly precarious [@Cornelius-BellBell2021].
This obviously poses a serious risk to the sustainability and growth of open source software in archaeology: if the people who occupy central positions in the network cease to be active, then it is likely that the overall network would fragment.
Assessing and mitigating this risk should be a high priority for future research in this area.

<!-- TODO: this should be backed up with a table or figure. The analysis seems to have got lost somewhere. Or just cut? -->
We also applied the same betweenness centrality algorithm on a subnetwork whose links are based only on issues and comments, and not code contributions. In this subnetwork, people who commit less code have higher betweenness scores. However, many of the people with high betweenness from the graph representing all contributions also appear here. These people who appear in both of these lists have a tendency to contribute as both committers and as commenters. This list also includes a series of contributors who never or rarely commit code. Although it is out of this study's scope a qualitative analysis of issues and comments may yield more insight on the kinds of contributions that each of these participants make.

>>>>

```{r fig-graph-repo-repo}
#| fig-cap: Graph of open archaeology repositories connected by common contributors. Darker edges indicate a great number of common contributors. Node colour indicates membership of the largest clusters according to the edge–betweenness method [@girvan2002]. Excludes repositories with only a single user.
g_repo_repo <- oarch_graph_repo_repo |>
  group_by(cluster) |>
  mutate(cluster_n = n()) |>
  ungroup() |>
  mutate(
    cluster_rank = dense_rank(desc(cluster_n)),
    cluster = if_else(cluster_rank <= 7, cluster, NA)
  ) |>
  ggraph(layout = "fr", niter = 5000, grid = "nogrid") +
  geom_edge_link(aes(alpha = n)) +
  geom_point_interactive(aes(x, y, colour = cluster, tooltip = repo)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  theme(panel.background = element_rect(fill = "white"))

@@

```{r fig-graph-user-user}
#| fig-cap: Graph of open archaeology users connected by contributions to common repositories. Darker edges indicate a great number of common repositories. Node colour indicates membership of the largest clusters according to the edge–betweenness method [@girvan2002]. Excludes users that have only contributed to a single repository.
g_user_user <- oarch_graph_user_user |>
  group_by(cluster) |>
  mutate(cluster_n = n()) |>
  ungroup() |>
  mutate(
    cluster_rank = dense_rank(desc(cluster_n)),
    cluster = if_else(cluster_rank <= 7, cluster, NA)
  ) |>
  ggraph(layout = "fr", niter = 5000, grid = "nogrid") +
  geom_edge_link(aes(alpha = n)) +
  geom_point_interactive(aes(x, y, colour = cluster, tooltip = user)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  theme(panel.background = element_rect(fill = "white"))

@@

Likewise, we applied the same edge-betweenness community detection method to identify clusters in the user-user network, which reveal collaborative groupings.
We again see a complementary primary core connected to several more peripheral clusters, which are internally-cohesive and exhibit few connections with other peripheral clusters.
The central core bridges all the peripheral clusters.
The central core is not uniform, and comprises several relatively discrete sub-clusters representing collaborative sub-communities.
While these sub-clusters are internally cohesive, they exhibit enough connections to other members of the central core so as to not be considered as separate or peripheral clusters.

@@

Despite being the minority language, Python packages tend to be more central than R packages, on average.
Python projects in archaeology tend to be large, highly collaborative, and either infrastructural or of interest to other fields, so they have disproportionately high centrality.
Our findings support the notion that there are significant development patterns when working with these two languages, and further analysis to qualify this observation is warranted.

@@

<!-- Predictors of user centrality -->
Centrality to the user–user graph is weakly predicted by a user's overall rate of activity, as measured by their total number of contributions [@fig-user-betweenness-contribs]. 
We did not collect demographic data on users that appear in our dataset, but based on our own knowledge of the community we can observe that those highly central to the network tend to be employed in (junior) academic positions, or in a few cases in cultural heritage authorities, rather than specifically as research software engineers.
Such positions tend to not to actively reward or encourage software development, at least not on a par with more traditional academic outputs [@baxter2012], and are increasingly precarious [@cornelius-bell2021].
This obviously poses a serious risk to the sustainability and growth of open source software in archaeology: if the people who occupy central positions in the network cease to be active, then it is likely that the overall network would fragment.
Assessing and mitigating this risk should be a high priority for future research in this area.

### 2024-02-02 analysis/_07-conclusion.qmd
Updates to the text and general touch ups.
Moved stuff on git and github to a new background subsection, moved data caveats from conclusion to methods, updated some interpretations in section 6, update interpretation based on cumulative growth figure.

Archaeology is a "magpie" discipline, in that it assembles tools and methods that originate from a variety of other disciplines and industries to better understand the past [@wylie1989; @nilssonstutz2018]. 
Moreover, archaeologists are accustomed to working on shoestring budgets, having to adapt their tools to suit specific, non-generalizable situations, and applying a DIY work ethic to create and co-create with others [@morgan2012;  @caraher2019; @batist2021]. 
The freedom to reuse, modify and extend code for a variety of purposes is a boon for the development of improvized tech stacks and workflows called up in contexts of archaeological fieldwork and analysis, and archaeologists share with open source software developers a pragmatic mindset with regards to planning collective work and evaluating outcomes [@scollar1999]. 
It is therefore unsurprising that archaeologists are prolific users and developers of open source software.

@@

To investigate these concerns, we operationalised open-source collaborative experiences as the use of certain features of git and GitHub visible to us in data from the GitHub API. 
This excludes a wide range of other potential practices, most obviously the use of other means of version-control and code-hosting (though as we demonstrate in <!-- encode the section --> Section 3, the git/GitHub combination is overwhelmingly the most popular). 
It also excludes collaboration through offline or private channels, or forms of collaboration we don't know about. 
We did not directly observe or interview archaeological software developers, though our conclusions do draw heavily from our experience as members of that community ourselves. 
Our earliest data is from XXX and our study can say little about collaborative software development in archaeology before this point, though we know there was a significant amount of it [@ducke2013 ; @whallon1972].

These caveats notwithstanding, we have documented that open source software development in archaeology has seen a rapid and sustained rise beginning around 2014. 
This is marked by a variety of applications and use cases, including the use of git and GitHub to track and host content other than code. 
Moreover, archaeologists are very involved in broader scripting ecosystems, as is evident through the predominant creation of R packages and Python libraries designed to process the rich variety of archaeological information. 
At the same time, archaeologists also create standalone software for more intensive tasks that require greater access to system resources or that warrant more complex user interfaces than what R and Python IDEs are capable of providing.

@@

<!-- This entire paragraph is subject to change based on revised network analysis -->
Our network analysis similarly draws attention to the real-world collaborative ties that underpin archaeological open source software development. We identified three clusters of close associations between projects: 
the first is centred around a highly specialized supdiscipline exhibiting a very enthusiastic attitude (**ZB NOTE: or maybe even pressure?**) toward working in the open; 
the second is concerned with developing solutions to a very common and discipline-wide problem, specifically relating to data collection and management in fieldwork; 
and the third deals with several topic of concern, but represents an emerging community of practice. 
This third cluster comprises efforts performed by members of a close-knit community, many of whom already collaborate at local research centres or participate in relatively private interest groups. 
This indicates that open source is firmly embedded within existing social and institutional support and power structures that permeate academic life, both online and offline.

There is thus little evidence to support the notion that archaeologists benefit from the positive outcomes that are commonly argued to be the natural results of open source development models -- namely, greater degrees of transparency, extensibility and participatory action.
While opening the source code may facilitate these positive outcomes as necessary preconditional factors, we argue that this only amounts to establishing the _potential_ for people to put these values into practice. 
Moreover, we argue that the objectives and circumstances that frame archaeological practice significantly influence how far archaeologists (and academics in general) are willing to push for these values, and limit the ability for archaeologists to do open source in ways that resemble more mainstream open source projects. 
For instance, successful open source projects like the Linux kernel or the Firefox web browser are driven by collective and popular interest in ensuring that code remains functional, and the code base is therefore constantly in flux and bears an accumulating list of contributing members; this significantly differs from the organizational principles that govern archaeological work, namely that a project's leaders runs the whole show from the top down and operationalize all other actors as instruments of their central directives. 
Moreover, archaeological projects ultimately seek to produce stable textual outcomes bearing clear delineation of authorship and that require no upkeep whatsoever. 
Sustaining an open source project is simply not compatible with the factors that currently drive the momentum behind archaeological work.

>>>>

@@

To investigate these concerns, we operationalised open-source collaborative experiences as the use of certain features of git and GitHub visible to us in data from the GitHub API. 
With this data, we documented that open source software development in archaeology has seen a rapid and sustained rise beginning around 2014. 
This is marked by a variety of applications and use cases, including the use of git and GitHub to track and host content other than code. 
Moreover, archaeologists are very involved in broader scripting ecosystems, as is evident through the predominant creation of R packages and Python libraries designed to process the rich variety of archaeological information. 
At the same time, archaeologists also create standalone software for more intensive tasks that require greater access to system resources or that warrant more complex user interfaces than what R and Python IDEs are capable of providing.

@@

Our network analysis similarly draws attention to the real-world collaborative ties that underpin archaeological open source software development.
We identify a core cluster representing a series of collaboratibe ties among members of a archaeological software developer community of practice.
This core exhibits complexity that correspond with social patterns.
Indeed, we demonstrate that real-world social connections and institutional support structures are strong predictors of centrality, with sub-clusters appearing within the central core mapping onto established partnerships.
This indicates that archaeological open source is firmly embedded within existing power structures that permeate academic life, both online and offline.

There is thus little evidence to support the notion that archaeologists benefit from the positive outcomes that are commonly argued to be the natural results of open source development models -- namely, greater degrees of transparency, extensibility and participatory action.
While opening the source code may facilitate these positive outcomes as necessary preconditional factors, we argue that this only amounts to establishing the _potential_ for people to put these values into practice. 
Moreover, we argue that the objectives and circumstances that frame archaeological practice significantly influence how far archaeologists (and academics in general) are willing to push for these values, and limit the ability for archaeologists to do open source in ways that resemble more mainstream open source projects.
For instance, successful open source projects like the Linux kernel or the Firefox web browser are driven by collective and popular interest in ensuring that code remains functional, and the code base is therefore constantly in flux and bears an accumulating list of contributing members.
This significantly differs from the organizational principles that govern archaeological work, namely that a project's leaders runs the whole show from the top down and operationalize all other actors as instruments of their central directives. 
Moreover, archaeological projects ultimately seek to produce stable textual outcomes bearing clear delineation of authorship and that require no upkeep whatsoever. 
Sustaining an open source project is simply not compatible with the factors that currently drive the momentum behind archaeological work.
